---
title: "Bikeshare"
author: "Pyimoe Than"
date: "6/17/2022"
output: word_document
---
##Business problem

How do annual members and casual riders use Cyclistic bikes differently?

## Importing library

```{r}
library(dplyr)
library(writexl)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(data.table)
library(lubridate)#for datetime object
```
## Load the data
```{r}
Bike_Share_202004=read.csv("202004-divvy-tripdata.csv",header=T)
Bike_Share_202005=read.csv("202005-divvy-tripdata.csv",header=T)
Bike_Share_202006=read.csv("202006-divvy-tripdata.csv",header=T)
Bike_Share_202007=read.csv("202007-divvy-tripdata.csv",header=T)
Bike_Share_202008=read.csv("202008-divvy-tripdata.csv",header=T)
Bike_Share_202009=read.csv("202009-divvy-tripdata.csv",header=T)
Bike_Share_202010=read.csv("202010-divvy-tripdata.csv",header=T)
Bike_Share_202011=read.csv("202011-divvy-tripdata.csv",header=T)
Bike_Share_202012=read.csv("202012-divvy-tripdata.csv",header=T)
```
## Total Number of Columns and Width

```{r}
dim(Bike_Share_202004)
dim(Bike_Share_202005)
dim(Bike_Share_202006)
dim(Bike_Share_202007)
dim(Bike_Share_202008)
dim(Bike_Share_202009)
dim(Bike_Share_202010)
dim(Bike_Share_202011)
dim(Bike_Share_202012)

```
## Check Column name of each dataset for consistency

```{r}
colnames(Bike_Share_202004)
colnames(Bike_Share_202005)
colnames(Bike_Share_202006)
colnames(Bike_Share_202007)
colnames(Bike_Share_202008)
colnames(Bike_Share_202009)
colnames(Bike_Share_202010)
colnames(Bike_Share_202011)
colnames(Bike_Share_202012)
```
## What kinds of Data we have in the dataset

```{r}
str(Bike_Share_202004)
str(Bike_Share_202005)
str(Bike_Share_202006)
str(Bike_Share_202007)
str(Bike_Share_202008)
str(Bike_Share_202009)
str(Bike_Share_202010)
str(Bike_Share_202011)
str(Bike_Share_202012)
```
## Convert numeric ID into Categorical ID

```{r}
Bike_Share_202004$start_station_id=as.character(Bike_Share_202004$start_station_id)
Bike_Share_202004$end_station_id=as.character(Bike_Share_202004$end_station_id)

Bike_Share_202005$start_station_id=as.character(Bike_Share_202005$start_station_id)
Bike_Share_202005$end_station_id=as.character(Bike_Share_202005$end_station_id)

Bike_Share_202006$start_station_id=as.character(Bike_Share_202006$start_station_id)
Bike_Share_202006$end_station_id=as.character(Bike_Share_202006$end_station_id)


Bike_Share_202007$start_station_id=as.character(Bike_Share_202007$start_station_id)
Bike_Share_202007$end_station_id=as.character(Bike_Share_202007$end_station_id)

Bike_Share_202008$start_station_id=as.character(Bike_Share_202008$start_station_id)
Bike_Share_202008$end_station_id=as.character(Bike_Share_202008$end_station_id)

Bike_Share_202009$start_station_id=as.character(Bike_Share_202009$start_station_id)
Bike_Share_202009$end_station_id=as.character(Bike_Share_202009$end_station_id)

Bike_Share_202010$start_station_id=as.character(Bike_Share_202010$start_station_id)
Bike_Share_202010$end_station_id=as.character(Bike_Share_202010$end_station_id)

Bike_Share_202011$start_station_id=as.character(Bike_Share_202011$start_station_id)
Bike_Share_202011$end_station_id=as.character(Bike_Share_202011$end_station_id)

Bike_Share_202012$start_station_id=as.character(Bike_Share_202012$start_station_id)
Bike_Share_202012$end_station_id=as.character(Bike_Share_202012$end_station_id)
```

## Combine all data and make it all trips
```{r}
Trips=rbind(Bike_Share_202004,Bike_Share_202005,Bike_Share_202006,Bike_Share_202007,Bike_Share_202008,Bike_Share_202009,Bike_Share_202010,Bike_Share_202011,Bike_Share_202012)
str(Trips)
```

## Check Missing Values

```{r}
#Margin=1 means check missing values in row
#Margin=2 means check missing values in column
apply(X=is.na(Trips),MARGIN=2,FUN=sum)
```
The data set has missing values in start_station_id column, end_station_id column, end_lat column and end_lng column.

## rename variable label

```{r}
Trips=Trips%>%
  rename(member_type=member_casual)
```

## Check duplicate data

```{r}
count(distinct(Trips))
```


Based on the above results, the dataset has no duplicate rows.

## Right now started_at column and ended_at column are in character format. It should be in datetime format. Convert those columns into datetime format.

```{r}
Trips$started_at=strptime(Trips$started_at,format="%Y-%m-%d %H: %M: %S")

Trips$ended_at=strptime(Trips$ended_at,format="%Y-%m-%d %H: %M: %S")
```

## Remove start_lat,start_lng, end_lat, and end_lng columns since those columns are not useful in our analysis

```{r}
Trips=Trips%>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
```
## Create the new column called ride_length in minutes

```{r}
Trips$ride_length=difftime(Trips$ended_at,Trips$started_at,units="mins")
```

## calculate the day of the week that each ride started

```{r}
Trips$day_of_week=wday(Trips$started_at,label=TRUE)
```
## calculate the month that each ride started

```{r}
Trips$month=month(Trips$started_at,label=TRUE)
```
## Checking for negative values in ride_length column. I Will remove those since it doesn't make sense to have a negative length of the ride

```{r}
Trips%>%count(ride_length<0)
Trips=Trips%>%filter(ride_length>0)
```

Remove 10548 rows since they have a negative values in ride_length column.

## Analyze Phase

## How many customers are casual and paid member

```{r}
Trips%>%
  group_by(member_type)%>%
  summarize(Num_Ride=n())
```
## Average Ride_length

```{r}
ave=Trips%>%
  group_by(member_type)%>%
  summarize(Average_Ride_length=mean(ride_length))
ave
```
On average, paid member rides the bike less minutes than casual member. This mean that casual members ride the bike longer duration than paid members. 


## In Which day do customers bike the most?And How long do they bike?

```{r}
Day=Trips%>%
  group_by(member_type,day_of_week)%>%
  select(member_type,day_of_week,ride_length)%>%
  summarize(Num_Rides=n(),average_ride=mean(ride_length))%>%
  arrange(member_type,day_of_week)
Day
```
## Data visualizations

```{r}
Trips%>%
  group_by(member_type,day_of_week)%>%
  summarize(Num_Rides=n())%>%
  ggplot(aes(x=day_of_week,y=Num_Rides,fill=member_type))+
  theme(axis.text.x=element_text(angle=30))+
  labs(title="Total Trips by customer in a week")+
  geom_col(width=0.5, position=position_dodge(width=0.5))+
  scale_y_continuous(labels=function(x) format(x,scientific=FALSE))
```

On Weekend, Casual and paid member ride the bike the most. From Monday to Friday, casual member ride decrease. However, paid member ride is still close to weekend ride. On Sunday, casual member ride an average of 51.73 minute. Casual member rides more duration than paid member.


## In which month do customers ride the most?

```{r}
month=Trips%>%
  group_by(member_type,month)%>%
  select(member_type,month,ride_length)%>%
  summarize(Num_Rides=n(),average_ride=mean(ride_length))%>%
  arrange(member_type,month)
month
```
## Data Visualizations of which month do customers ride the most.

```{r}
Trips%>%
  group_by(member_type,month)%>%
  summarize(Num_Rides=n())%>%
  ggplot(aes(x=month,y=Num_Rides,fill=member_type))+
  theme(axis.text.x=element_text(angle=30))+
  labs(title="Total Trips by customer in each month")+
  geom_col(width=0.5, position=position_dodge(width=0.5))+
  scale_y_continuous(labels=function(x) format(x,scientific=FALSE))
```
During the peak of summer months, casual and paid member ride the most. After the summer season is over, number of rides for casual and paid member decrease significantly. And also, average number of rides also decrease after the summer is over.

## What kinds of ride do customer like the most?

```{r}
Type=Trips%>%
  group_by(member_type,rideable_type)%>%
  select(member_type,rideable_type,ride_length)%>%
  summarize(Num_Rides=n(),average_ride=mean(ride_length))%>%
  arrange(member_type,-Num_Rides)
Type
```
## Data visualizations for what kinds of ride do customer like the most.

```{r}
Trips%>%
  group_by(member_type,rideable_type)%>%
  summarize(Num_Rides=n())%>%
  ggplot(aes(x=rideable_type,y=Num_Rides,fill=member_type))+
  labs(title="Type of ride vs. Number of Rides")+
  theme(axis.text.x=element_text(angle=30))+
  geom_col(width=0.5, position=position_dodge(width=0.5))+
  scale_y_continuous(labels=function(x) format(x,scientific=FALSE))
```
Docked Bikes are the most popular type of bike for both casual and paid member. Both customers don't usually use classic bike. 
#```{r}
#write.csv(Trips,'C:\\Users\\Aungkyaw\\Desktop\\data.csv')
#```